---
title: "CellScore"
author: "Sophie.Alice"
date: "2026-07-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Setup

```{r, message=FALSE, warning = FALSE}
rm(list=ls())
gc()

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(hdf5r)
library(scCustomize)
library(patchwork)
library(cowplot)
library(ggrepel)
library(workflowr)
library(scDblFinder)
library(clustree)
library(harmony)
library(speckle)
library(MAST)
library(tibble)
library(EnhancedVolcano)

library(ggarchery)
library(ggtangle)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(DOSE)
library(fgsea)
library(forcats)
library(SCpubr)
library(assertthat)
library(UCell)
library(DT)

library(ggforce) 
library(enrichplot)
library(ggrain)
library(ComplexHeatmap)
library(tidyverse)
library(circlize)
library(effsize)
library(ggpubr)

```

# Get the Data

```{r, warning = FALSE}
path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/Post_Annotation.rds"
Data <- read_rds(path)
Data

Fibroblasts <- subset(Data, cell_type == "Cardiac Fibroblasts")
Idents(Fibroblasts) <- "orig.ident" 

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
gradient_colors3 <- custom_palette(6)
gradient_colors4 <- custom_palette(4)
gradient_colors2 <- custom_palette(2)
```


# Functions

```{r, warning = FALSE}



#################################################### Raincloud Plots

raincloud.plots <- function(gene_list, 
                                   plot_title, 
                                   plot_subtitle, 
                                   seurat_obj = Data, 
                                   colors = gradient_colors4) {

  
  valid_genes <- intersect(gene_list, rownames(seurat_obj))
  expression_matrix <- FetchData(seurat_obj, vars = c(valid_genes, "ident")) 

  
  rain_data <- expression_matrix %>%
    tibble::rownames_to_column("Cell_ID") %>%
    tidyr::pivot_longer(
      cols = all_of(valid_genes), 
      names_to = "Gene", 
      values_to = "Expression"
    ) %>%
    dplyr::rename(Condition = ident)
  

  raincloud_plot <- ggplot(rain_data, aes(x = Condition, y = Expression, fill = Condition, color = Condition)) +
    geom_rain(
      alpha = 0.6,
      violin.args = list(color = NA),          
      boxplot.args = list(outlier.shape = NA), 
      point.args = list(
        size = 0.4,                            
        alpha = 0.4                            
      ),
    group = 1
    ) +
    
    facet_wrap(~Gene, scales = "free_y", ncol = 4) +
    
    scale_fill_manual(values = colors) +
    scale_color_manual(values = colors) +
    
    theme_bw() +
    labs(
      title    = plot_title,
      subtitle = plot_subtitle,
      x        = "Experimental Conditions",
      y        = "Normalized Expression Level"
    ) +
    
    theme(
      plot.title.position = "plot",
      plot.title    = element_text(face = "bold", size = 12, hjust = 0),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0),
      axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
      axis.text.y   = element_text(face = "bold", color = "black"),
      axis.title.x  = element_text(face = "bold"),
      axis.title.y  = element_text(face = "bold"),
      strip.text    = element_text(face = "bold", size = 10), 
      legend.position = "none"
    )
  
  return(raincloud_plot)
}


```



# Main Cellscores

```{r, warning = FALSE}
pro_fibrotic_genes <- c(
  "ADAM12", "ADAM19", "ADAMTS2", "ADAMTS4", "ADAMTS6", "ADAMTS12", 
  "ADAMTS14", "ADAMTS16", "ADAMTS17", 
  "CCN2", "CDH2", "CDH11", 
  "COL4A1", "COL4A2", "COL5A1", "COL5A2", "COL8A1", "COL8A2", 
  "COL10A1", "COL11A1", "COL12A1", "COL13A1", "COL15A1", 
  "COMP", "FAP", "FLRT2", "FN1", "IGFBP7", "ITGA11", "ITGAV", "ITGB1", 
  "ITGB5", "ITGBL1", "LAMB1", "MMP13", "PDGFA", "PLAU", "PLAUR", 
  "PLOD2", "POSTN", "PXDN", "RHOA", "RUNX1", "SEMA7A", "SERPINE1", 
  "SPHK1", "SPOCK1", "SRPX2", "TGFBR1", "THBS1", "THBS2", "THY1", 
  "TIMP3", "TNFAIP6", "TNFRSF12A", "VIM"
)

Fibroblasts <- AddModuleScore_UCell(
  obj = Fibroblasts,
  features = list(Pro_Fibrotic = pro_fibrotic_genes),
  name = "_UCell"
)

######################################

rescue.score <- c(
  "MMP2", "VEGFA", "RCN3", "SRPX2", "ADAM12", "PXDN", "LOXL3", "FBN1", 
  "HMGA2", "THBS1", "SERPINE1", "HAS2", "GREM1", "WNT5A", "SEMA3C", 
  "TNFRSF12A", "ITGB1", "FGF2", "COL8A1", "XBP1", "LRP1", "COL12A1", 
  "POSTN", "ADAMTS6", "ADAMTSL1", "COL14A1", "FBLN5", "ADAMTS1", "ELN")

Fibroblasts <- AddModuleScore_UCell(
  obj = Fibroblasts,
  features = list(rescue_score = rescue.score),
  name = "_UCell"
)

```

# Pro-Fibrotic score comparison

```{r, warning = FALSE, fig.width=8, fig.height=6}

df_tgf <- Fibroblasts@meta.data %>% 
  dplyr::filter(orig.ident %in% c("CMF_TGF", "CMFHM_TGF"))

# Define comparison group
my_comparisons <- list(c("CMF_TGF", "CMFHM_TGF"))

# color for the 2 conditions
colors <- c("CMF_TGF" = "#005EA8", "CMFHM_TGF" = "#9F1C67")

# Plot UCell
p_tgf <- ggplot(df_tgf, aes(x = orig.ident, y = Pro_Fibrotic_UCell, fill = orig.ident)) +
  geom_violin(alpha = 0.5, color = NA, scale = "area") +
  geom_boxplot(width = 0.2, color = "black", outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = colors) +
  
  # Add pairwise Wilcoxon test p-value
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.signif" 
  ) +
  
  theme_bw() +
  labs(
    title    = "Pro-Fibrotic Gene Score: TGF-β Response ± Macrophages",
    subtitle = "CMF_TGF (No Macrophages) vs CMFHM_TGF (With Macrophages)",
    x        = "Condition",
    y        = "UCell Score"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title    = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle = element_text(face = "italic", size = 10, hjust = 0),
    axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y   = element_text(face = "bold", color = "black"),
    axis.title.x  = element_text(face = "bold"),
    axis.title.y  = element_text(face = "bold"),
    legend.position = "none"
  )

print(p_tgf)


FibroblastsTGF <- subset(Fibroblasts, TGF_status == "TGF_B")
table(FibroblastsTGF$orig.ident)

##########################################

p_split <- FeaturePlot_scCustom(
  FibroblastsTGF, 
  features = "Pro_Fibrotic_UCell", 
  split.by = "orig.ident"
) 

p_split <- p_split + 
  plot_annotation(
    title    = "Pro-Fibrotic Gene Score: TGF-β Response ± Macrophages",
    subtitle = "CMF_TGF (No Macrophages) vs CMFHM_TGF (With Macrophages)",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 12, hjust = 0),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0)
    )
  ) & 
  theme(
    aspect.ratio = 1.4,
    axis.text.y  = element_text(size = 12, hjust = 0),
    axis.title   = element_text(face = "plain"))

print(p_split)


```

# Adding Cliffs delta

The Cliffs delta to show effect size : 

it measures how often values in one group are larger than in another (ranging from -1 to 1)

Interpretation Scale for Cliff's Delta (d):

 - $\vert{}d\vert{} < 0.147$: Negligible
 - $\vert{}d\vert{} < 0.33$: Small
 - $\vert{}d\vert{} < 0.474$: Medium
 - $\vert{}d\vert{} \ge 0.474$: Large


Results: 

delta estimate: 0.4221959 (medium)

a medium d in combination with sign. p suggests that the pro-fibrotic score is decreased when macrophages are present.!

```{r, warning = FALSE, fig.width=8, fig.height=6}

# Calculate Cliff's Delta between CMF_TGF and CMFHM_TGF
cliff.delta(Pro_Fibrotic_UCell ~ orig.ident, data = df_tgf)

```

## Checking the individual gene scores 

```{r, warning = FALSE, fig.width=8, fig.height=6}

chunk_size <- 25
total_genes <- length(pro_fibrotic_genes)

if (total_genes > 0) {
  for (i in seq(1, total_genes, by = chunk_size)) {
    
    end_idx <- min(i + chunk_size - 1, total_genes)
    slice_num <- ((i - 1) / chunk_size) + 1

    print(SCpubr::do_DotPlot(
      sample = FibroblastsTGF, 
      features = pro_fibrotic_genes[i:end_idx],
      dot.scale = 10,
      plot.title = "Pro-fibrotic Cell score: TGF-β Response ± Macrophages",
      plot.subtitle = paste0("Slice ", slice_num, " (Genes ", i, " to ", end_idx, " of ", total_genes, ")")
    ))
  }
}


```

```{r, warning = FALSE, fig.width=8, fig.height=6}

chunk_size <- 12
total_genes <- length(pro_fibrotic_genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = pro_fibrotic_genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs CMF_TGF: Gene distribution profiles across conditions",
    plot_subtitle = paste0("Geneset = pro-fibrotic score - Slice ", slice_num),
    seurat_obj    = FibroblastsTGF,
    colors        = gradient_colors2
  ))
}

```

# Pro-Fibrotic Module score separated

```{r, warning = FALSE, fig.width=8, fig.height=6}


Fibroblasts <- subset(Data, cell_type == "Cardiac Fibroblasts")
Idents(Fibroblasts) <- "orig.ident" 

fibrosis_modules_expanded <- list(
  ECM_Structural = c(
    "COL4A1", "COL4A2", "COL5A1", "COL5A2", "COL8A1", "COL8A2", 
    "COL10A1", "COL11A1", "COL12A1", "COL13A1", "COL15A1", 
    "FN1", "COMP", "POSTN", "LAMB1", "VIM"),
  ECM_Remodeling = c(
    "ADAM12", "ADAM19", "ADAMTS2", "ADAMTS4", "ADAMTS6", "ADAMTS12", 
    "ADAMTS14", "ADAMTS16", "ADAMTS17", "MMP13", "PLAU", "PLAUR", 
    "SERPINE1", "TIMP3"),
  ECM_Crosslinking = c(
    "PLOD2", "PXDN", "SPOCK1", "TNFAIP6"),
  Cell_Adhesion = c(
    "ITGA11", "ITGAV", "ITGB1", "ITGB5", "ITGBL1", 
    "CDH2", "CDH11", "FAP", "THY1", "FLRT2", "RHOA"),
  Fibrotic_Signaling = c(
    "CCN2", "PDGFA", "TGFBR1", "THBS1", "THBS2", "SEMA7A", 
    "TNFRSF12A", "IGFBP7", "SPHK1", "SRPX2", "RUNX1"
  )
)

Fibroblasts <- AddModuleScore_UCell(
  obj = Fibroblasts,
  features = fibrosis_modules_expanded,
  name = "_UCell"
)

```


```{r, warning = FALSE, fig.width=8, fig.height=6}
Fibroblasts$cardiomyocyte.genes_UCell <- NULL # removing this so it doesn't interfere with my plot

my_comparisons <- list(c("CMF_TGF", "CMFHM_TGF"))

df_long <- Fibroblasts@meta.data %>% 
  dplyr::filter(orig.ident %in% c("CMF_TGF", "CMFHM_TGF")) %>%
  pivot_longer(
    cols = ends_with("_UCell"),
    names_to = "Module",
    values_to = "Score"
  ) %>%
  mutate(
    Module = gsub("_UCell", "", Module),
    Module = gsub("_", " ", Module)
  )

colors <- c("CMF_TGF" = "#005EA8", "CMFHM_TGF" = "#9F1C67")

# Plot facetted comp
p_modules <- ggplot(df_long, aes(x = orig.ident, y = Score, fill = orig.ident, color = orig.ident)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_jitter(width = 0.15, size = 0.3, alpha = 0.2) +
  geom_boxplot(width = 0.2, color = "black", outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  facet_wrap(~ Module, scales = "free_y", ncol = 2) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme_bw() +
  labs(
    title    = "Deconstruction of Pro-Fibrotic Response ± Macrophages",
    subtitle = "UCell functional sub-modules (CMF_TGF vs CMFHM_TGF)",
    x        = "Condition",
    y        = "UCell Score"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title       = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle    = element_text(face = "italic", size = 10, hjust = 0),
    strip.background = element_rect(fill = "grey92", color = NA),
    strip.text       = element_text(face = "bold", size = 10),
    axis.text.x      = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y      = element_text(face = "bold", color = "black"),
    legend.position  = "none"
  )

print(p_modules)

```


# Rescue score 

```{r, warning = FALSE, fig.width=8, fig.height=6}


# Plot UCell
p_tgf <- ggplot(df_tgf, aes(x = orig.ident, y = rescue_score_UCell, fill = orig.ident)) +
  geom_violin(alpha = 0.5, color = NA, scale = "area") +
  geom_boxplot(width = 0.2, color = "black", outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = colors) +
  
  # Add pairwise Wilcoxon test p-value
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +

  theme_bw() +
  labs(
    title    = "Rescue Score: TGF-β Response ± Macrophages",
    subtitle = "CMF_TGF (No Macrophages) vs CMFHM_TGF (With Macrophages)",
    x        = "Condition",
    y        = "UCell Score"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title    = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle = element_text(face = "italic", size = 10, hjust = 0),
    axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y   = element_text(face = "bold", color = "black"),
    axis.title.x  = element_text(face = "bold"),
    axis.title.y  = element_text(face = "bold"),
    legend.position = "none"
  )

print(p_tgf)

p_split <- FeaturePlot_scCustom(
  FibroblastsTGF, 
  features = "rescue_score_UCell", 
  split.by = "orig.ident"
) 

p_split <- p_split + 
  plot_annotation(
    title    = "Rescue Score: TGF-β Response ± Macrophages",
    subtitle = "CMF_TGF (No Macrophages) vs CMFHM_TGF (With Macrophages)",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 12, hjust = 0),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0)
    )
  ) & 
  theme(
    aspect.ratio = 1.4,
    axis.text.y  = element_text(size = 12, hjust = 0),
    axis.title   = element_text(face = "plain"))

print(p_split)

```

# Adding Cliffs delta

Results: 

delta estimate: 0.7087503 (large)

a large d in combination with sign. p suggests that the pro-fibrotic score is decreased when macrophages are present.!


```{r, warning = FALSE, fig.width=8, fig.height=6}

# Calculate Cliff's Delta between CMF_TGF and CMFHM_TGF
cliff.delta(rescue_score_UCell ~ orig.ident, data = df_tgf)

```

# Subsetted module scores 

```{r, warning = FALSE, fig.width=8, fig.height=6}

Fibroblasts <- subset(Data, cell_type == "Cardiac Fibroblasts")
Idents(Fibroblasts) <- "orig.ident" 

fibrosis_modules <- list(
  ECM_Deposition = c("COL8A1", "COL12A1", "COL14A1", "ELN", "LOXL3", "PXDN", "HAS2", "FBN1", "FBLN5", "POSTN"),
  ECM_Remodeling = c("MMP2", "ADAM12", "ADAMTS1", "ADAMTS6", "ADAMTSL1", "SERPINE1"),
  Cell_Matrix_Adhesion = c("ITGB1", "LRP1", "TNFRSF12A", "XBP1"),
  Fibrotic_Signaling = c("THBS1", "GREM1", "WNT5A", "VEGFA", "FGF2", "SEMA3C", "SRPX2", "RCN3", "HMGA2")
)

Fibroblasts <- AddModuleScore_UCell(
  obj = Fibroblasts,
  features = fibrosis_modules,
  name = "_UCell"
)

```

```{r, warning = FALSE, fig.width=8, fig.height=6}
Fibroblasts$cardiomyocyte.genes_UCell <- NULL # removing this so it doesn't interfere with my plot

my_comparisons <- list(c("CMF_TGF", "CMFHM_TGF"))


df_long <- Fibroblasts@meta.data %>% 
  dplyr::filter(orig.ident %in% c("CMF_TGF", "CMFHM_TGF")) %>%
  pivot_longer(
    cols = ends_with("_UCell"),
    names_to = "Module",
    values_to = "Score"
  ) %>%
  mutate(
    Module = gsub("_UCell", "", Module),
    Module = gsub("_", " ", Module)
  )

colors <- c("CMF_TGF" = "#005EA8", "CMFHM_TGF" = "#9F1C67")

# Plot facetted comp
p_modules <- ggplot(df_long, aes(x = orig.ident, y = Score, fill = orig.ident, color = orig.ident)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_jitter(width = 0.15, size = 0.3, alpha = 0.2) +
  geom_boxplot(width = 0.2, color = "black", outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  facet_wrap(~ Module, scales = "free_y", ncol = 2) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  
  stat_compare_means(
    comparisons = my_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme_bw() +
  labs(
    title    = "Deconstruction of Rescue Score ± Macrophages",
    subtitle = "UCell functional sub-modules (CMF_TGF vs CMFHM_TGF)",
    x        = "Condition",
    y        = "UCell Score"
  ) +
  theme(
    plot.title.position = "plot",
    plot.title       = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle    = element_text(face = "italic", size = 10, hjust = 0),
    strip.background = element_rect(fill = "grey92", color = NA),
    strip.text       = element_text(face = "bold", size = 10),
    axis.text.x      = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
    axis.text.y      = element_text(face = "bold", color = "black"),
    legend.position  = "none"
  )

print(p_modules)

```

```{r, warning = FALSE, fig.width=8, fig.height=6}


```

```{r, warning = FALSE, fig.width=8, fig.height=6}


```

```{r, warning = FALSE, fig.width=8, fig.height=6}


```

```{r, warning = FALSE, fig.width=8, fig.height=6}


```

```{r, warning = FALSE, fig.width=8, fig.height=6}


```
