---
title: "CM Subtypes"
author: "Sophie.Alice"
date: "2026-06-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---


# Setup

```{r, message=FALSE, warning = FALSE}
rm(list=ls())
gc()

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(hdf5r)
library(scCustomize)
library(patchwork)
library(cowplot)
library(ggrepel)
library(workflowr)
library(scDblFinder)
library(clustree)
library(harmony)
library(speckle)
library(MAST)
library(tibble)
library(EnhancedVolcano)

library(ggarchery)
library(ggtangle)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(DOSE)
library(fgsea)
library(forcats)
library(SCpubr)
library(assertthat)
library(UCell)
library(DT)
```

# Get the Data

```{r, warning = FALSE}
# path <- "/shares/rheumatologie.usz/Sophie/Gabi_Projects/Laila_MT_Project/RDS_Files/Post_QC.rds"
path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/Post_Annotation.rds"
Data <- read_rds(path)
Data

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
gradient_colors3 <- custom_palette(6)
gradient_colors4 <- custom_palette(4)

DimPlot(Data, reduction = "umap", shuffle = T,
        group.by = "SCT_snn_res.0.1", label=T)
```

# Subset only the CM

```{r, warning = FALSE}

Data <- subset(Data, !(cell_type %in% c("Cardiac Fibroblasts", "Macrophages")))

DimPlot(Data, reduction = "umap", shuffle = T,
        group.by = "SCT_snn_res.0.1", label=T)

```

# Functions

```{r, warning = FALSE}


Volcano_Plots <- function(Data, Title, Subtitle, pCutoff = 0.05, FCcutoff = 0.5) {
  
  # Clean extreme values
  df <- Data
  df <- df[is.finite(df$avg_log2FC) & is.finite(df$p_val_adj), ]
  
  if (any(df$p_val_adj == 0)) {
    min_nonzero <- min(df$p_val_adj[df$p_val_adj > 0], na.rm = TRUE)

    if(is.infinite(min_nonzero)) min_nonzero <- 1e-500 
    df$p_val_adj[df$p_val_adj == 0] <- min_nonzero
  }
  
  # dynamic axis limits
  xmax <- max(df$avg_log2FC, na.rm = TRUE)
  xmin <- min(df$avg_log2FC, na.rm = TRUE)
  ymax <- max(-log10(df$p_val_adj), na.rm = TRUE)
  
  # Add padding
  xpad <- (xmax - xmin) * 0.1
  ypad <- ymax * 0.1
  
  combined_subtitle <- paste0(Subtitle, " ● Thresholds: FDR < ", pCutoff, ", |Log2FC| > ", FCcutoff)
  
  EV <- EnhancedVolcano(
    df,
    lab = df$gene,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    xlab = 'log2 Foldchange',
    ylab = "-log10 FDR",
    legendLabels = c('NS','log2FC','FDR','FDR & log2FC'),
    col = c("darkgrey", "#005EA8", "#FFD966", "#9F1C67"),
    title = Title,
    subtitle = combined_subtitle,   
    xlim = c(xmin - xpad, xmax + xpad),
    ylim = c(0, ymax + ypad),
    pCutoff = pCutoff,              
    FCcutoff = FCcutoff,            

    cutoffLineType = 'dashed',      
    cutoffLineCol = 'black',        
    cutoffLineWidth = 0.6,
    labSize = 3.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5
  )
  
  return(EV)
}

############################## Datatables 

create_table <- function(df) {
  DT::datatable(
    df,
    extensions = 'Buttons', 
    options = list(
      dom = 'Blfrtip',                       
      buttons = c('copy', 'csv', 'excel', 'pdf'),
      pageLength = 10,                      
      lengthMenu = c(10, 25, 50, 100)     
    ),
    filter = 'top',                         
    class = 'cell-border stripe hover'       
  )
}

```


# Difference between Cluster 0 and Cluste 4 (Hypoxia?)

## DGEA

```{r, warning = FALSE}

deg_list <- list()

table(Data$SCT_snn_res.0.1)
Idents(Data) <- "SCT_snn_res.0.1"

MAST.Res <- FindMarkers(
  object = Data,
  ident.1 = "4",    
  ident.2 = "0",     
  test.use = "MAST",
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)

create_table(MAST.Res.sign)

deg_list[["0 vs. 4"]] <- MAST.Res


Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG Cluster 0 vs. Cluster 4",
  Subtitle = "",
  pCutoff = 0.05,    
  FCcutoff = 0.5     
)


```

## FGSEA

### Get Hallmark pathways

```{r,warning = FALSE}
Hallmark.pathways <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/h.all.v2026.1.Hs.symbols.gmt")
```

### Rank the data

```{r,warning = FALSE}

DEG <- deg_list[["0 vs. 4"]]

DEG.ranked <- DEG %>%
  dplyr::select(gene, avg_log2FC) %>%
  na.omit() %>%
  distinct() %>%
  group_by(gene) %>%
  summarize(LFC = mean(avg_log2FC)) %>%
  arrange(desc(LFC))

ranked <- deframe(DEG.ranked)

rank <- ranked[names(ranked) != ""]

```

### Functional gene set enrichment Analysis

```{r,warning = FALSE}

############################## Hallmark
set.seed(42)
fgseares <- fgsea(pathways = Hallmark.pathways, stats = rank, nperm = 1000)

fgseares.tidy <- fgseares %>%
  as_tibble() %>%  
  arrange(desc(NES))

fgseares.tidy <- fgseares.tidy %>%
  dplyr::select(-leadingEdge ,-ES, -nMoreExtreme) 

fgseares.tidy.top.up   <- head(fgseares.tidy, 15)
fgseares.tidy.top.down <- tail(fgseares.tidy, 15)
fgseares.tidy.top.both <- rbind(fgseares.tidy.top.up, fgseares.tidy.top.down)

create_table(fgseares.tidy.top.both)

```

### Visualization

```{r,warning = FALSE}

plot_df <- fgseares.tidy.top.both %>%
  mutate(
    padj_group = case_when(
      padj <= 0.05 ~ "< 0.05",
      padj <= 0.2  ~ "0.05 – 0.2",
      TRUE         ~ "> 0.2"
    ),
    padj_group = factor(padj_group, levels = c("< 0.05", "0.05 – 0.2", "> 0.2")),
    pathway = gsub("HALLMARK_", "", pathway),
  )


ggplot(plot_df, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = padj_group), color = "black", linewidth = 0.2) +
  coord_flip() +
  scale_x_discrete(position = "top", labels = scales::label_wrap(50)) +
  scale_fill_manual(
    name = "adj. p-value",
    values = c(
      "< 0.05"     = "#9F1C67",   
      "0.05 – 0.2" = "#FFD966",   
      "> 0.2"      = "#005EA8"    
    ),
    drop = FALSE 
  ) +
  labs(
    x = "", 
    y = "Normalized Enrichment Score (NES)", 
    title = "GSEA: Hallmark Pathways",
    subtitle = "MT: Pathway changes Cluster 4 compared to Cluster 0"
  ) +
  theme_bw() +
  theme(
    aspect.ratio    = 1.4, 
    axis.text.y     = element_text(size = 8, hjust = 0), 
    axis.title      = element_text(face = "bold"),
    plot.title      = element_text(hjust = 0, face = "bold"),
    plot.subtitle   = element_text(hjust = 0, size = 9, face = "italic"),
    legend.position = "left" 
  )


```

## PAH

```{r,warning = FALSE}

df <- deg_list[["0 vs. 4"]]

# Upregulated Genes
up_genes <- df %>%
  filter(avg_log2FC > 0.5, p_val_adj <= 0.05) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

# Downregulated Genes
down_genes <- df %>%
  filter(avg_log2FC < -0.5, p_val_adj <= 0.05) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

# Background Universe Genes 
universe <- df %>%
  distinct(gene) %>% 
  pull(gene) %>%                        
  na.omit()

```


```{r, warning = FALSE}

set.seed(42)

Humanlibrary <- org.Hs.eg.db

###################################### upregulated genes

up.GO <- enrichGO(
  gene          = up_genes,
  universe      = universe,         
  OrgDb         = Humanlibrary,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,       
  minGSSize     = 15,
  maxGSSize     = 800
)

simp.up.GO <- clusterProfiler::simplify(up.GO)

simp.up.GO.show <- as.data.frame(simp.up.GO)
simp.up.GO.show <- simp.up.GO.show %>%
  dplyr::select(-geneID) 
create_table(simp.up.GO.show)

###################################### downregulated genes

down.GO <- enrichGO(
  gene          = down_genes,
  universe      = universe,         
  OrgDb         = Humanlibrary,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,       
  minGSSize     = 15,
  maxGSSize     = 800
)

simp.down.GO <- clusterProfiler::simplify(down.GO)

simp.down.GO.show <- as.data.frame(simp.down.GO)
simp.down.GO.show <- simp.down.GO.show %>%
  dplyr::select(-geneID) 

create_table(simp.down.GO.show)

```

```{r, warning = FALSE}

go_up_df   <- as.data.frame(simp.up.GO) %>% mutate(Direction = "Up-regulated")
go_down_df <- as.data.frame(simp.down.GO) %>% mutate(Direction = "Down-regulated")

go_combined <- rbind(go_up_df, go_down_df) %>%
  group_by(Direction) %>%
  slice_head(n = 10) %>% # Adjust 'n' to show more or fewer terms per panel
  ungroup()


go_plot_data <- go_combined %>%
  mutate(
    GeneRatio_num = sapply(GeneRatio, function(x) {
      num_den <- as.numeric(strsplit(x, "/")[[1]])
      return(num_den[1] / num_den[2])})) %>%
  arrange(Direction, GeneRatio_num) %>%
  mutate(Description = factor(Description, levels = unique(Description)))


ggplot(go_plot_data, aes(x = GeneRatio_num, y = Description)) + 
  geom_point(aes(color = p.adjust, size = Count)) +  
  facet_wrap(~Direction, scales = "free_y", ncol = 1) +
  
  scale_color_viridis_c(
    option = "plasma", 
    name = "adj. p-value",                          
    guide = guide_colorbar(reverse = TRUE)
  ) +
  scale_size_continuous(range = c(3, 8), name = "Gene Count") +
  scale_y_discrete(labels = scales::label_wrap(45)) +
  
  theme_bw() + 
  labs(
    title = "DEG MT 0 vs 4: GO Biological Process Enrichment", 
    subtitle = "Comparing top enriched terms for Up-regulated vs Down-regulated genes",
    x = "Gene Ratio", 
    y = ""
  ) +
  theme(
    plot.title.position = "plot", 
    plot.title    = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle = element_text(size = 10, face = "italic", hjust = 0),
    axis.text.y   = element_text(size = 9, color = "black", face = "bold"),
    axis.title.x  = element_text(face = "bold"),
    strip.text    = element_text(face = "bold", size = 10), # Customizes facet header boxes
    legend.position = "right"
  )
```

## Check for Hypoxia markers

```{r, warning = FALSE}

# 3 different Hypoxia genesets

# M13324
hypox1 <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/BIOCARTA_HIF_PATHWAY.v2026.1.Hs.gmt")

# M5202
hypox2 <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/BIOCARTA_P53HYPOXIA_PATHWAY.v2026.1.Hs.gmt")

# M25398
hypox3 <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY_IN_RESPONSE_TO_HYPOXIA.v2026.1.Hs.gmt")

SCpubr::do_DotPlot(Data, 
                   features = hypox1,
                   dot.scale = 10,
                   plot.title = "Hypoxia Marker Expression Across Clusters") 

Data <- AddModuleScore_UCell(
  Data,
  features = hypox1
)

FeaturePlot_scCustom(Data, features = "BIOCARTA_HIF_PATHWAY_UCell")

#####################################################################

SCpubr::do_DotPlot(Data, 
                   features = hypox2,
                   dot.scale = 10,
                   plot.title = "Hypoxia Marker Expression Across Clusters") 

Data <- AddModuleScore_UCell(
  Data,
  features = hypox2
)

FeaturePlot_scCustom(Data, features = "BIOCARTA_P53HYPOXIA_PATHWAY_UCell")

#####################################################################

SCpubr::do_DotPlot(Data, 
                   features = hypox3,
                   dot.scale = 10,
                   plot.title = "Hypoxia Marker Expression Across Clusters") 

Data <- AddModuleScore_UCell(
  Data,
  features = hypox3
)

FeaturePlot_scCustom(Data, features = "GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY_IN_RESPONSE_TO_HYPOXIA_UCell")

```

# Conclusion Cluster 4

Highly propable that this cluster is the ischemic core of the MT.


# Difference between Cluster 0 and Cluste 5 (Unknown?)

## DGEA

```{r, warning = FALSE}

deg_list <- list()

table(Data$SCT_snn_res.0.1)
Idents(Data) <- "SCT_snn_res.0.1"

MAST.Res <- FindMarkers(
  object = Data,
  ident.1 = "5",    
  ident.2 = "0",     
  test.use = "MAST",
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)

create_table(MAST.Res.sign)

deg_list[["0 vs. 5"]] <- MAST.Res


Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG Cluster 0 vs. Cluster 5",
  Subtitle = "",
  pCutoff = 0.05,    
  FCcutoff = 0.5     
)


```

## FGSEA

### Rank the data

```{r,warning = FALSE}

DEG <- deg_list[["0 vs. 5"]]

DEG.ranked <- DEG %>%
  dplyr::select(gene, avg_log2FC) %>%
  na.omit() %>%
  distinct() %>%
  group_by(gene) %>%
  summarize(LFC = mean(avg_log2FC)) %>%
  arrange(desc(LFC))

ranked <- deframe(DEG.ranked)

rank <- ranked[names(ranked) != ""]

```

### Functional gene set enrichment Analysis

```{r,warning = FALSE}

############################## Hallmark
set.seed(42)
fgseares <- fgsea(pathways = Hallmark.pathways, stats = rank, nperm = 1000)

fgseares.tidy <- fgseares %>%
  as_tibble() %>%  
  arrange(desc(NES))

fgseares.tidy <- fgseares.tidy %>%
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) 


fgseares.tidy.top.up   <- head(fgseares.tidy, 15)
fgseares.tidy.top.down <- tail(fgseares.tidy, 15)
fgseares.tidy.top.both <- rbind(fgseares.tidy.top.up, fgseares.tidy.top.down)

create_table(fgseares.tidy.top.both)

# size is always 1 so this is not informative  
```

### Visualization

```{r,warning = FALSE}

plot_df <- fgseares.tidy.top.both %>%
  mutate(
    padj_group = case_when(
      padj <= 0.05 ~ "< 0.05",
      padj <= 0.2  ~ "0.05 – 0.2",
      TRUE         ~ "> 0.2"
    ),
    padj_group = factor(padj_group, levels = c("< 0.05", "0.05 – 0.2", "> 0.2")),
    pathway = gsub("HALLMARK_", "", pathway),
  )


ggplot(plot_df, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = padj_group), color = "black", linewidth = 0.2) +
  coord_flip() +
  scale_x_discrete(position = "top", labels = scales::label_wrap(50)) +
  scale_fill_manual(
    name = "adj. p-value",
    values = c(
      "< 0.05"     = "#9F1C67",   
      "0.05 – 0.2" = "#FFD966",   
      "> 0.2"      = "#005EA8"    
    ),
    drop = FALSE 
  ) +
  labs(
    x = "", 
    y = "Normalized Enrichment Score (NES)", 
    title = "GSEA: Hallmark Pathways",
    subtitle = "MT: Pathway changes Cluster 5 compared to Cluster 0"
  ) +
  theme_bw() +
  theme(
    aspect.ratio    = 1.4, 
    axis.text.y     = element_text(size = 8, hjust = 0), 
    axis.title      = element_text(face = "bold"),
    plot.title      = element_text(hjust = 0, face = "bold"),
    plot.subtitle   = element_text(hjust = 0, size = 9, face = "italic"),
    legend.position = "left" 
  )


```

## PAH

-> 0 enriched terms found

```{r,warning = FALSE}

df <- deg_list[["0 vs. 5"]]

# Upregulated Genes
up_genes <- df %>%
  filter(avg_log2FC > 0.5, p_val_adj <= 0.05) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

# Downregulated Genes
 # -> none

# Background Universe Genes 
universe <- df %>%
  distinct(gene) %>% 
  pull(gene) %>%                        
  na.omit()

```


```{r, warning = FALSE}

set.seed(42)

Humanlibrary <- org.Hs.eg.db

###################################### upregulated genes

up.GO <- enrichGO(
  gene          = up_genes,
  universe      = universe,         
  OrgDb         = Humanlibrary,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,       
  minGSSize     = 15,
  maxGSSize     = 800
)

up.GO


###################################### downregulated genes

# none present

```

### Look at the marker genes

```{r, warning = FALSE}
marker.gene.df <- deg_list[["0 vs. 5"]]
marker.genes <- marker.gene.df$gene
marker.genes.sign <- marker.gene.df %>%
  filter(avg_log2FC > 0.5, p_val_adj <= 0.05) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

FeaturePlot_scCustom(Data, features = marker.genes.sign[1:6])
FeaturePlot_scCustom(Data, features = marker.genes.sign[7:12])
FeaturePlot_scCustom(Data, features = marker.genes.sign[13:16])

```


### First Slice of the Marker genes is the most pronounced: 

- **ENSG00000289426** = Antisense to **ANKRD45**
- **CLEC2A** = also regulation of actin filament polymerization (GeneCards); facilitates immune recognition of keratinocytes leading to natural killer cell mediated cytotoxicity
- **ANKRD45** = The protein is linked to proliferative control and to structures that organize the final stages of cell division (GeneCards)
- **ENSG00000289578** = Antisense to **FAM3C** (Also upregulated)
- **KLHL20**: Promotes Lys-33-linked ubiquitination and proteasomal degradation; Enhances **HIF-1** Signaling
- **KLRF2**: Is activated by **CLEC2A** 
- **ENSG00000286339**: Antsense to **CEP85L**
- **SBF2**: Guanine nucleotide exchange factor (GEF) which activates **RAB21** and possibly **RAB28** (key regulators of intracellular membrane trafficking)
- **CEP85L**: centrosome-associated protein 
- **FAM3C**: linked to signal transduction and gene-expression programs. Extracellular regulator rather than an intracellular enzyme.?


# Conlucion for Cluster 5

??

