---
title: "DGEA"
author: "Sophie.Alice"
date: "2026-06-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Setup

```{r, message=FALSE}
rm(list=ls())
gc()

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(hdf5r)
library(scCustomize)
library(patchwork)
library(cowplot)
library(ggrepel)
library(workflowr)
library(scDblFinder)
library(clustree)
library(harmony)
library(speckle)
library(MAST)
library(tibble)
library(EnhancedVolcano)
library(UpSetR)
library(grid)
library(ggvenn)

#BiocManager::install("MAST")
```

# Get the Data

```{r}
# path <- "/shares/rheumatologie.usz/Sophie/Gabi_Projects/Laila_MT_Project/RDS_Files/Post_QC.rds"
path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/Post_Annotation.rds"
Data <- read_rds(path)
Data

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
gradient_colors3 <- custom_palette(6)
gradient_colors4 <- custom_palette(4)
```

# Functions

```{r}


Volcano_Plots <- function(Data, Title, Subtitle, pCutoff = 0.05, FCcutoff = 0.5) {
  
  # Clean extreme values
  df <- Data
  df <- df[is.finite(df$avg_log2FC) & is.finite(df$p_val_adj), ]
  
  # dynamic axis limits
  xmax <- max(df$avg_log2FC, na.rm = TRUE)
  xmin <- min(df$avg_log2FC, na.rm = TRUE)
  ymax <- max(-log10(df$p_val_adj), na.rm = TRUE)
  
  # Add padding
  xpad <- (xmax - xmin) * 0.1
  ypad <- ymax * 0.1
  
  combined_subtitle <- paste0(Subtitle, " ● Thresholds: FDR < ", pCutoff, ", |Log2FC| > ", FCcutoff)
  
  EV <- EnhancedVolcano(
    df,
    lab = NA,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    xlab = 'log2 Foldchange',
    ylab = "-log10 FDR",
    legendLabels = c('NS','log2FC','FDR','FDR & log2FC'),
    col = c("darkgrey", "#005EA8", "#FFD966", "#9F1C67"),
    title = Title,
    subtitle = combined_subtitle,   
    xlim = c(xmin - xpad, xmax + xpad),
    ylim = c(0, ymax + ypad),
    pCutoff = pCutoff,              
    FCcutoff = FCcutoff,            

    cutoffLineType = 'dashed',      
    cutoffLineCol = 'black',        
    cutoffLineWidth = 0.6           
  )
  
  return(EV)
}


```

# Fibroblasts

## Subset the data

```{r}

table(Data$cell_type)
table(Data$orig.ident)

Idents(Data) <- "orig.ident" 
Fibroblasts<- subset(Data, cell_type == "Cardiac Fibroblasts")

Fibroblasts$cdr <- scale(colSums(GetAssayData(Fibroblasts, assay = "RNA", layer = "counts") > 0))


fibroblast_deg_list <- list()

```

## 1. Baseline Fibrotic response to TGF-B without Macrophages

```{r}


MAST.Res <- FindMarkers(
  object = Fibroblasts,
  ident.1 = "CMF_TGF",    
  ident.2 = "CMF_NC",     
  test.use = "MAST",
  latent.vars = "cdr",      
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)



fibroblast_deg_list[["Baseline_TGF_Response"]] <- MAST.Res.sign

head(MAST.Res.sign)


Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG in MT-Fibroblasts without Macrophages",
  Subtitle = "TGF-\u03b2 vs. Control",
  pCutoff = 0.01,    
  FCcutoff = 1     
)

#write.csv(MAST.Res, file = file.path(outputdir, "Baseline_TGF_Response_all.csv"))
#write.csv(MAST.Res.sign, file = file.path(outputdir, "Baseline_TGF_Response_sign.csv"))

```

## 2. Modulated Fibrotic response to TGF-B with Macrophages

```{r}


MAST.Res <- FindMarkers(
  object = Fibroblasts,
  ident.1 = "CMFHM_TGF",    
  ident.2 = "CMFHM_NC",     
  test.use = "MAST",
  latent.vars = "cdr",      
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)

head(MAST.Res.sign)

fibroblast_deg_list[["Modulated_TGF_Response"]] <- MAST.Res.sign


Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG in MT-Fibroblasts with Macrophages",
  Subtitle = "TGF-\u03b2 vs. Control",
  pCutoff = 0.01,    
  FCcutoff = 1     
)

#write.csv(MAST.Res, file = file.path(outputdir, "Modulated_TGF_Response_all.csv"))
#write.csv(MAST.Res.sign, file = file.path(outputdir, "Modulated_TGF_Response_sign.csv"))

```

## 3. Baseline Macrophage response

```{r}


MAST.Res <- FindMarkers(
  object = Fibroblasts,
  ident.1 = "CMFHM_NC",    
  ident.2 = "CMF_NC",     
  test.use = "MAST",
  latent.vars = "cdr",      
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)

head(MAST.Res.sign)

fibroblast_deg_list[["Baseline_Mac_Response"]] <- MAST.Res.sign

Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG in MT-Fibroblasts unstimulated",
  Subtitle = "with Macrophages vs. Control",
  pCutoff = 0.01,    
  FCcutoff = 1     
)

#write.csv(MAST.Res, file = file.path(outputdir, "Baseline_Mac_Response_all.csv"))
#write.csv(MAST.Res.sign, file = file.path(outputdir, "Baseline_Mac_Response_sign.csv"))

```

## 4. Stimulated Macrophage Effect

```{r}

MAST.Res <- FindMarkers(
  object = Fibroblasts,
  ident.1 = "CMFHM_TGF",    
  ident.2 = "CMF_TGF",     
  test.use = "MAST",
  latent.vars = "cdr",      
  logfc.threshold = 0.1,      
  min.pct = 0.2             
)

MAST.Res <- MAST.Res %>%
  rownames_to_column(var = "gene")

MAST.Res.sign <- MAST.Res %>%
  filter(p_val_adj < 0.05 & abs(avg_log2FC) >= 0.5)

head(MAST.Res.sign)

fibroblast_deg_list[["Stim_Mac_Response"]] <- MAST.Res.sign


Volcano_Plots(
  Data = MAST.Res,
  Title = "DEG in MT-Fibroblasts stimulated with TGF-\u03b2",
  Subtitle = "with Macrophages vs. Control",
  pCutoff = 0.01,    
  FCcutoff = 1     
)

#write.csv(MAST.Res, file = file.path(outputdir, "Stim_Mac_Response_all.csv"))
#write.csv(MAST.Res.sign, file = file.path(outputdir, "Stim_Mac_Response_sign.csv"))

```

# Gene Crossover

```{r}

upset_up_genes   <- list()
upset_down_genes <- list()


p_cutoff  <- 0.01
fc_cutoff <- 1 

for (comp_name in names(fibroblast_deg_list)) {

  df <- fibroblast_deg_list[[comp_name]]
  
  if (!"gene" %in% colnames(df)) {
    df <- df %>% rownames_to_column(var = "gene")
  }
  
  upset_up_genes[[comp_name]] <- df %>%
    filter(p_val_adj <= p_cutoff & avg_log2FC >= fc_cutoff) %>%
    pull(gene) %>%
    na.omit()
  
  upset_down_genes[[comp_name]] <- df %>%
    filter(p_val_adj <= p_cutoff & avg_log2FC <= -fc_cutoff) %>%
    pull(gene) %>%
    na.omit()
}

# one master object
upset_final_data <- list(
  Up_Regulated   = upset_up_genes,
  Down_Regulated = upset_down_genes
)


```


```{r}

################################################### Upregulated Intersections

upset(
  fromList(upset_final_data$Up_Regulated), 
  nsets = 4,                       
  order.by = "freq",               
  main.bar.color = "#9F1C67",      
  sets.bar.color = "darkgrey",     
  text.scale = c(1.3, 1.3, 1, 1, 1.2, 1) 
)


################################################### Downregulated Intersections

upset(
  fromList(upset_final_data$Down_Regulated), 
  nsets = 4, 
  order.by = "freq", 
  main.bar.color = "#005EA8", 
  sets.bar.color = "darkgrey",
  text.scale = c(1.3, 1.3, 1, 1, 1.2, 1)
)

```
# Venn diagram of all ovelapping genes

```{r}

################################# upregulated genes

ggvenn(
  data = upset_final_data$Up_Regulated, 
  columns = names(upset_final_data$Up_Regulated),                          
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Upregulated Genes",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 16, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

################################# downregulated genes

ggvenn(
  data = upset_final_data$Down_Regulated, 
  columns = names(upset_final_data$Down_Regulated),                          
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Downregulated Genes",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 16, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )


```

# Venn diagram of DEG upon TGF-B stimulation with and without Macrophages

```{r}

################################# upregulated genes

ggvenn(
  data = upset_final_data$Up_Regulated[c("Baseline_TGF_Response", "Modulated_TGF_Response")], 
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Upregulated Genes upon TGF-\u03b2 stimulation",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

################################# downregulated genes

ggvenn(
  data = upset_final_data$Down_Regulated[c("Baseline_TGF_Response", "Modulated_TGF_Response")], 
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Downregulated Genes upon TGF-\u03b2 stimulation",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )


```


# Venn diagram of DEG of presence of Macrophages with and without TGF-B stimulation

```{r}
################################# upregulated genes

ggvenn(
  data = upset_final_data$Up_Regulated[c("Baseline_Mac_Response", "Stim_Mac_Response")],
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Upregulated Genes upon Macrophage presence",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

################################# downregulated genes

ggvenn(
  data = upset_final_data$Down_Regulated[c("Baseline_Mac_Response", "Stim_Mac_Response")],  
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Downregulated Genes upon Macrophage presence",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 13, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

```


```{r}

```
