---
title: "Fibroblasts_DEG_Overlap"
author: "Sophie.Alice"
date: "2026-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---


# Setup

```{r, message=FALSE, warning = FALSE}
rm(list=ls())
gc()

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(hdf5r)
library(scCustomize)
library(patchwork)
library(cowplot)
library(ggrepel)
library(workflowr)
library(scDblFinder)
library(clustree)
library(harmony)
library(speckle)
library(MAST)
library(tibble)
library(EnhancedVolcano)
library(UpSetR)
library(grid)
library(ggvenn)

library(ggarchery)
library(ggtangle)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(DOSE)
library(fgsea)
library(forcats)
library(SCpubr)
library(assertthat)
library(UCell)
library(DT)

library(ggforce) 
library(enrichplot)
library(ggrain)
```


# Get the Data

```{r, warning = FALSE}
path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/Post_Annotation.rds"
Data <- read_rds(path)
Data
Fibroblasts<- subset(Data, cell_type == "Cardiac Fibroblasts")
Idents(Fibroblasts) <- "orig.ident" 

path2 <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/All_Fibro_DEG_Lists.rds"

DEG.lists <- read_rds(path2)

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
gradient_colors3 <- custom_palette(6)
gradient_colors4 <- custom_palette(4)
gradient_colors3 <- custom_palette(3)
```


# Functions

```{r, warning = FALSE}


Volcano_Plots <- function(Data, Title, Subtitle, pCutoff = 0.05, FCcutoff = 0.5) {
  
  # Clean extreme values
  df <- Data
  df <- df[is.finite(df$avg_log2FC) & is.finite(df$p_val_adj), ]
  
  # dynamic axis limits
  xmax <- max(df$avg_log2FC, na.rm = TRUE)
  xmin <- min(df$avg_log2FC, na.rm = TRUE)
  ymax <- max(-log10(df$p_val_adj), na.rm = TRUE)
  
  # Add padding
  xpad <- (xmax - xmin) * 0.1
  ypad <- ymax * 0.1
  
  combined_subtitle <- paste0(Subtitle, " ● Thresholds: FDR < ", pCutoff, ", |Log2FC| > ", FCcutoff)
  
  EV <- EnhancedVolcano(
    df,
    lab = df$gene,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    xlab = 'log2 Foldchange',
    ylab = "-log10 FDR",
    legendLabels = c('NS','log2FC','FDR','FDR & log2FC'),
    col = c("darkgrey", "#005EA8", "#FFD966", "#9F1C67"),
    title = Title,
    subtitle = combined_subtitle,   
    xlim = c(xmin - xpad, xmax + xpad),
    ylim = c(0, ymax + ypad),
    pCutoff = pCutoff,              
    FCcutoff = FCcutoff,            

    cutoffLineType = 'dashed',      
    cutoffLineCol = 'black',        
    cutoffLineWidth = 0.6,
    labSize = 3.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5
  )
  
  return(EV)
}

##################################### Tables

create_table <- function(df) {
  DT::datatable(
    df,
    extensions = 'Buttons', 
    options = list(
      dom = 'Blfrtip',                       
      buttons = c('copy', 'csv', 'excel', 'pdf'),
      pageLength = 10,                      
      lengthMenu = c(10, 25, 50, 100)     
    ),
    filter = 'top',                         
    class = 'cell-border stripe hover'       
  )
}

#################################################### Raincloud Plots

raincloud.plots <- function(gene_list, 
                                   plot_title, 
                                   plot_subtitle, 
                                   seurat_obj = Data, 
                                   colors = gradient_colors4) {

  
  valid_genes <- intersect(gene_list, rownames(seurat_obj))
  expression_matrix <- FetchData(seurat_obj, vars = c(valid_genes, "ident")) 

  
  rain_data <- expression_matrix %>%
    tibble::rownames_to_column("Cell_ID") %>%
    tidyr::pivot_longer(
      cols = all_of(valid_genes), 
      names_to = "Gene", 
      values_to = "Expression"
    ) %>%
    dplyr::rename(Condition = ident)
  

  raincloud_plot <- ggplot(rain_data, aes(x = Condition, y = Expression, fill = Condition, color = Condition)) +
    geom_rain(
      alpha = 0.6,
      violin.args = list(color = NA),          
      boxplot.args = list(outlier.shape = NA), 
      point.args = list(
        size = 0.4,                            
        alpha = 0.4                            
      ),
    group = 1
    ) +
    
    facet_wrap(~Gene, scales = "free_y", ncol = 3) +
    
    scale_fill_manual(values = colors) +
    scale_color_manual(values = colors) +
    
    theme_bw() +
    labs(
      title    = plot_title,
      subtitle = plot_subtitle,
      x        = "Experimental Conditions",
      y        = "Normalized Expression Level"
    ) +
    
    theme(
      plot.title.position = "plot",
      plot.title    = element_text(face = "bold", size = 12, hjust = 0),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0),
      axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
      axis.text.y   = element_text(face = "bold", color = "black"),
      axis.title.x  = element_text(face = "bold"),
      axis.title.y  = element_text(face = "bold"),
      strip.text    = element_text(face = "bold", size = 10), 
      legend.position = "none"
    )
  
  return(raincloud_plot)
}


```

# Gene Overlap

## Get all the DEG

Cutoffs Chosen: FDR <= 0.01, abs(log2FC) > 1

```{r, warning = FALSE}

names(DEG.lists) <- c("Baseline_TGF_Response" = "CMF_TGF vs. CMF_NC",
                      "Modulated_TGF_Response" = "CMFHM_TGF vs. CMFHM_NC",
                      "Baseline_Mac_Response" = "CMFHM_NC vs. CMF_NC",
                      "Stim_Mac_Response" = "CMFHM_TGF vs. CMF_TGF")

# head(DEG.lists)

```



```{r, warning = FALSE}

upset_up_genes   <- list()
upset_down_genes <- list()


p_cutoff  <- 0.01
fc_cutoff <- 1

for (comp_name in names(DEG.lists)) {

  df <- DEG.lists[[comp_name]]
  
  if (!"gene" %in% colnames(df)) {
    df <- df %>% rownames_to_column(var = "gene")
  }
  
  upset_up_genes[[comp_name]] <- df %>%
    filter(p_val_adj <= p_cutoff & avg_log2FC >= fc_cutoff) %>%
    pull(gene) %>%
    na.omit()
  
  upset_down_genes[[comp_name]] <- df %>%
    filter(p_val_adj <= p_cutoff & avg_log2FC <= -fc_cutoff) %>%
    pull(gene) %>%
    na.omit()
}

# one master object
upset_final_data <- list(
  Up_Regulated   = upset_up_genes,
  Down_Regulated = upset_down_genes
)


```

### UpsetPlot for all Groups

```{r, warning = FALSE}

################################################### Upregulated Intersections

upset(
  fromList(upset_final_data$Up_Regulated), 
  nsets = 4,                       
  order.by = "freq",               
  main.bar.color = "#9F1C67",      
  sets.bar.color = "darkgrey",     
  text.scale = c(1.3, 1.3, 1, 1, 1.2, 1) 
)


################################################### Downregulated Intersections

upset(
  fromList(upset_final_data$Down_Regulated), 
  nsets = 4, 
  order.by = "freq", 
  main.bar.color = "#005EA8", 
  sets.bar.color = "darkgrey",
  text.scale = c(1.3, 1.3, 1, 1, 1.2, 1)
)

```

### Venn diagram of all ovelapping genes

```{r, warning = FALSE}

################################# upregulated genes

ggvenn(
  data = upset_final_data$Up_Regulated, 
  columns = names(upset_final_data$Up_Regulated),                          
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Upregulated Genes",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 16, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

################################# downregulated genes

ggvenn(
  data = upset_final_data$Down_Regulated, 
  columns = names(upset_final_data$Down_Regulated),                          
  fill_color = gradient_colors4, 
  stroke_size = 0.5, 
  set_name_size = 4
) +
  labs(
    title = "Ovelap of sign. Downregulated Genes",
    subtitle = paste0("Fibroblast Differential Expression ● Cutoffs: FDR < ", p_cutoff, ", |Log2FC| > ", fc_cutoff)
  ) +
  theme(
    plot.title = element_text(hjust = 0, face = "bold", size = 16, color = "black"),
    plot.subtitle = element_text(hjust = 0, face = "italic", size = 11, color = "dimgray")
  )

```


## Log2FC differences across Codnitions

```{r, warning = FALSE}

library(ComplexHeatmap)
library(tidyverse)
library(circlize)

FDR <- 0.01 

# Extract gene lists and filter only for significance
processed_lists <- lapply(names(DEG.lists), function(name) {
  df <- DEG.lists[[name]]
  df %>%
    filter(p_val_adj < FDR) %>%
    dplyr::select(gene, avg_log2FC) %>%
    rename(!!name := avg_log2FC) 
})

merged_df <- purrr::reduce(processed_lists, full_join, by = "gene")

# Convert NAs to 0 (meaning 0 log2FC change when not significant)
merged_df[is.na(merged_df)] <- 0

fc_matrix <- as.matrix(merged_df[,-1])
rownames(fc_matrix) <- merged_df$gene


if (nrow(fc_matrix) > 200) {
  top_genes <- head(order(apply(fc_matrix, 1, var), decreasing = TRUE), 200)
  fc_matrix_subset <- fc_matrix[top_genes, ]
} else {
  fc_matrix_subset <- fc_matrix
}

max_val <- max(abs(fc_matrix_subset))
col_fun = colorRamp2(c(-max_val, 0, max_val), c("#005EA8", "white", "#9F1C67"))


Heatmap(
  fc_matrix_subset, 
  name = "log2FC", 
  col = col_fun,
  
  cluster_columns = FALSE,        
  cluster_rows = TRUE,            
  show_row_dend = TRUE,           
  
  column_names_rot = 45,          
  column_names_side = "bottom",   
  show_row_names = FALSE,          
  row_names_gp = gpar(fontsize = 8), 
  
  row_km = 4,                     
  column_title = "log2FC Differences across Conditions"
)

```

## Check on genes that are up regulated in CMF_TGF vs. CMF_NC AND down regulated in CMFHM_TGF vs. CMF_TGF 

 - Display of the log2FC across Contitions
 
```{r, warning = FALSE}

FDR <- 0.01 

filtered_df <- merged_df[merged_df$`CMF_TGF vs. CMF_NC` > 0.5 & merged_df$`CMFHM_TGF vs. CMF_TGF` < -0.5, ]

filtered_df.genes <- filtered_df$gene

length(filtered_df.genes)


fc_matrix <- as.matrix(filtered_df[,-1])
rownames(fc_matrix) <- filtered_df$gene


max_val <- max(abs(fc_matrix))
col_fun = colorRamp2(c(-max_val,0, max_val), c("#005EA8","white", "#9F1C67"))


Heatmap(
  fc_matrix, 
  name = "log2FC", 
  col = col_fun,
    column_title = "log2FC Changes of Genes of Interest\nlog2FC CMF_TGF vs. CMF_NC > 0.5 & CMFHM_TGF vs. CMF_TGF < -0.5", 
  column_title_gp = gpar(
    fontsize = 13, 
    col = "black"
  ),
  cluster_columns = FALSE,        
  cluster_rows = TRUE,            
  show_row_dend = TRUE,           
  
  column_names_rot = 45,          
  column_names_side = "bottom",   
  show_row_names = FALSE,          
  row_names_gp = gpar(fontsize = 8), 
  
  row_km = 4
)
```

## Check on genes that are up regulated in CMF_TGF vs. CMF_NC AND down regulated in CMFHM_TGF vs. CMF_TGF 

 - Display of the scaled relative gene expressions of the Genes of Interest

```{r, warning = FALSE, fig.width=8, fig.height=12}

Idents(Fibroblasts) <- "orig.ident"
avg_matrix <- AverageExpression(Fibroblasts, assays = "SCT", features = filtered_df.genes)$SCT


scaled_matrix <- t(scale(t(avg_matrix)))
scaled_matrix.df <- as.data.frame(scaled_matrix)
create_table(scaled_matrix.df)


col_fun = colorRamp2(c(-2, 0, 2), c("#005EA8","white", "#9F1C67")) 

H1 <- Heatmap(
  scaled_matrix, 
  name = "Z-score", 
  col = col_fun,
  
  column_title = "Relative Gene Expression of Genes of Interest\nlog2FC CMF_TGF vs. CMF_NC > 0.5 & CMFHM_TGF vs. CMF_TGF < -0.5", 
  column_title_gp = gpar(
    fontsize = 13, 
    col = "black"
  ),
  
  cluster_columns = FALSE,         
  cluster_rows = TRUE,             
  show_row_dend = TRUE,            
  
  column_names_rot = 45,          
  column_names_side = "bottom",   
  show_row_names = TRUE,          
  row_names_gp = gpar(fontsize = 6), 
  row_km = 4
)

print(H1)
```


## Checking on the 4 Blocks

```{r, warning = FALSE, fig.width=8, fig.height=12}

ht_shifted <- draw(H1)

m_blocks <- row_order(ht_shifted)

# get gene namese per block

block1_genes <- rownames(scaled_matrix)[m_blocks[[1]]]
block2_genes <- rownames(scaled_matrix)[m_blocks[[2]]]
block3_genes <- rownames(scaled_matrix)[m_blocks[[3]]]
block4_genes <- rownames(scaled_matrix)[m_blocks[[4]]]

```




### Block 1

```{r, warning = FALSE, fig.width=8, fig.height=5}

block1_genes 

SCpubr::do_DotPlot(Fibroblasts, 
                   features = block1_genes,
                   dot.scale = 10,
                   plot.title = "avg. Gene Expression of G.o.I - Block 1") 

chunk_size <- 9
total_genes <- length(block1_genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = block1_genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
    plot_subtitle = paste0("Geneset = positive regulation of cell migration - Slice ", slice_num),
    seurat_obj    = Fibroblasts,
    colors        = gradient_colors4
  ))
}

```
