---
title: "Modulated_TGF_Response Lenient Thresholds"
author: "Sophie.Alice"
date: "2026-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---


# Setup

```{r, message=FALSE, warning = FALSE}
rm(list=ls())
gc()

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(hdf5r)
library(scCustomize)
library(patchwork)
library(cowplot)
library(ggrepel)
library(workflowr)
library(scDblFinder)
library(clustree)
library(harmony)
library(speckle)
library(MAST)
library(tibble)
library(EnhancedVolcano)

library(ggarchery)
library(ggtangle)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(DOSE)
library(fgsea)
library(forcats)
library(SCpubr)
library(assertthat)
library(UCell)
library(DT)

library(ggforce) 
library(enrichplot)
library(ggrain)
```


# Get the Data

```{r, warning = FALSE}
path <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/Post_Annotation.rds"
Data <- read_rds(path)
Data
Fibroblasts<- subset(Data, cell_type == "Cardiac Fibroblasts")
Idents(Fibroblasts) <- "orig.ident" 

path2 <- "/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/RDS_Files/All_Fibro_DEG_Lists.rds"

DEG.lists <- read_rds(path2)

custom_palette <- colorRampPalette(c("#005EA8", "#6DB335", "#FFD966", "#EDAAD2", "#9F1C67"))
gradient_colors3 <- custom_palette(6)
gradient_colors4 <- custom_palette(4)
```


# Functions

```{r, warning = FALSE}


Volcano_Plots <- function(Data, Title, Subtitle, pCutoff = 0.05, FCcutoff = 0.5) {
  
  # Clean extreme values
  df <- Data
  df <- df[is.finite(df$avg_log2FC) & is.finite(df$p_val_adj), ]
  
  # dynamic axis limits
  xmax <- max(df$avg_log2FC, na.rm = TRUE)
  xmin <- min(df$avg_log2FC, na.rm = TRUE)
  ymax <- max(-log10(df$p_val_adj), na.rm = TRUE)
  
  # Add padding
  xpad <- (xmax - xmin) * 0.1
  ypad <- ymax * 0.1
  
  combined_subtitle <- paste0(Subtitle, " ● Thresholds: FDR < ", pCutoff, ", |Log2FC| > ", FCcutoff)
  
  EV <- EnhancedVolcano(
    df,
    lab = df$gene,
    x = 'avg_log2FC',
    y = 'p_val_adj',
    xlab = 'log2 Foldchange',
    ylab = "-log10 FDR",
    legendLabels = c('NS','log2FC','FDR','FDR & log2FC'),
    col = c("darkgrey", "#005EA8", "#FFD966", "#9F1C67"),
    title = Title,
    subtitle = combined_subtitle,   
    xlim = c(xmin - xpad, xmax + xpad),
    ylim = c(0, ymax + ypad),
    pCutoff = pCutoff,              
    FCcutoff = FCcutoff,            

    cutoffLineType = 'dashed',      
    cutoffLineCol = 'black',        
    cutoffLineWidth = 0.6,
    labSize = 3.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5
  )
  
  return(EV)
}

##################################### Tables

create_table <- function(df) {
  DT::datatable(
    df,
    extensions = 'Buttons', 
    options = list(
      dom = 'Blfrtip',                       
      buttons = c('copy', 'csv', 'excel', 'pdf'),
      pageLength = 10,                      
      lengthMenu = c(10, 25, 50, 100)     
    ),
    filter = 'top',                         
    class = 'cell-border stripe hover'       
  )
}

#################################################### Raincloud Plots

raincloud.plots <- function(gene_list, 
                                   plot_title, 
                                   plot_subtitle, 
                                   seurat_obj = Data, 
                                   colors = gradient_colors4) {

  
  valid_genes <- intersect(gene_list, rownames(seurat_obj))
  expression_matrix <- FetchData(seurat_obj, vars = c(valid_genes, "ident")) 

  
  rain_data <- expression_matrix %>%
    tibble::rownames_to_column("Cell_ID") %>%
    tidyr::pivot_longer(
      cols = all_of(valid_genes), 
      names_to = "Gene", 
      values_to = "Expression"
    ) %>%
    dplyr::rename(Condition = ident)
  

  raincloud_plot <- ggplot(rain_data, aes(x = Condition, y = Expression, fill = Condition, color = Condition)) +
    geom_rain(
      alpha = 0.6,
      violin.args = list(color = NA),          
      boxplot.args = list(outlier.shape = NA), 
      point.args = list(
        size = 0.4,                            
        alpha = 0.4                            
      ),
    group = 1
    ) +
    
    facet_wrap(~Gene, scales = "free_y", ncol = 3) +
    
    scale_fill_manual(values = colors) +
    scale_color_manual(values = colors) +
    
    theme_bw() +
    labs(
      title    = plot_title,
      subtitle = plot_subtitle,
      x        = "Experimental Conditions",
      y        = "Normalized Expression Level"
    ) +
    
    theme(
      plot.title.position = "plot",
      plot.title    = element_text(face = "bold", size = 12, hjust = 0),
      plot.subtitle = element_text(face = "italic", size = 10, hjust = 0),
      axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold", color = "black"),
      axis.text.y   = element_text(face = "bold", color = "black"),
      axis.title.x  = element_text(face = "bold"),
      axis.title.y  = element_text(face = "bold"),
      strip.text    = element_text(face = "bold", size = 10), 
      legend.position = "none"
    )
  
  return(raincloud_plot)
}


```

# DEGs of Modulated Fibrotic response to TGF-B with Macrophages

 - CMFHM_TGF vs CMFHM_NC = Effect of TGF-Beta on MT with Macrophages

 - Cutoffs Chosen: FDR <= 0.05, abs(log2FC) > 0.5


```{r, warning = FALSE, fig.width=8, fig.height=7}


FDR <- 0.05
FC <- 0.5

Degs <- DEG.lists$Modulated_TGF_Response

Degs.sign <- Degs %>%
  filter(p_val_adj < FDR & abs(avg_log2FC) >= FC)

create_table(Degs.sign)

Volcano_Plots(
  Data = Degs,
  Title = "DEG in MT-Fibroblasts with Macrophages",
  Subtitle = "TGF-\u03b2 vs. Control",
  pCutoff = FDR,    
  FCcutoff = FC     
)

```

# FGSEA

## Get Hallmark pathways

```{r,warning = FALSE}
Hallmark.pathways <- gmtPathways("/Users/sophiewagner/Desktop/USZ 25:26/Gabi_Project_Setup/Laila_MT_Project/Laila_MT_Project/Hallmark/h.all.v2026.1.Hs.symbols.gmt")
```

## Rank the data

```{r,warning = FALSE}

DEG.ranked <- Degs %>%
  dplyr::select(gene, avg_log2FC) %>%
  na.omit() %>%
  distinct() %>%
  group_by(gene) %>%
  summarize(LFC = mean(avg_log2FC)) %>%
  arrange(desc(LFC))

ranked <- deframe(DEG.ranked)

rank <- ranked[names(ranked) != ""]

```

## Functional gene set enrichment Analysis

```{r,warning = FALSE}

############################## Hallmark
set.seed(42)

fgseares <- fgsea(pathways = Hallmark.pathways, stats = rank, nperm = 1000)

fgseares.tidy <- fgseares %>%
  as_tibble() %>%  
  arrange(desc(NES))

fgseares.tidy <- fgseares.tidy %>%
  dplyr::select(-ES, -nMoreExtreme) 

fgseares.tidy.top.up   <- head(fgseares.tidy, 15)
fgseares.tidy.top.down <- tail(fgseares.tidy, 15)
fgseares.tidy.top.both <- rbind(fgseares.tidy.top.up, fgseares.tidy.top.down)

fgseares.tidy.show <- fgseares.tidy %>%
  dplyr::select(-leadingEdge) 
create_table(fgseares.tidy.show)

```

## Visualization

```{r,warning = FALSE, fig.width=6, fig.height=5}

plot_df <- fgseares.tidy.top.both %>%
  mutate(
    padj_group = case_when(
      padj <= 0.05 ~ "< 0.05",
      padj <= 0.2  ~ "0.05 – 0.2",
      TRUE         ~ "> 0.2"
    ),
    padj_group = factor(padj_group, levels = c("< 0.05", "0.05 – 0.2", "> 0.2")),
    pathway = gsub("HALLMARK_", "", pathway),
  )


ggplot(plot_df, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = padj_group), color = "black", linewidth = 0.2) +
  coord_flip() +
  scale_x_discrete(position = "top", labels = scales::label_wrap(50)) +
  scale_fill_manual(
    name = "adj. p-value",
    values = c(
      "< 0.05"     = "#9F1C67",   
      "0.05 – 0.2" = "#FFD966",   
      "> 0.2"      = "#005EA8"    
    ),
    drop = FALSE 
  ) +
  labs(
    x = "", 
    y = "Normalized Enrichment Score (NES)", 
    title = "GSEA: Hallmark Pathways",
    subtitle = "MT: Pathway changes CMFHM_TGF compared to CMFHM_NC"
  ) +
  theme_bw() +
  theme(
    aspect.ratio    = 1.4, 
    axis.text.y     = element_text(size = 8, hjust = 0), 
    axis.title      = element_text(face = "bold"),
    plot.title      = element_text(hjust = 0, face = "bold"),
    plot.subtitle   = element_text(hjust = 0, size = 9, face = "italic"),
    legend.position = "left" 
  )


```

# PAH

```{r,warning = FALSE}

df <- Degs

# Upregulated Genes
up_genes <- df %>%
  filter(avg_log2FC > FC, p_val_adj <= FDR) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

# Downregulated Genes
down_genes <- df %>%
  filter(avg_log2FC < -FC, p_val_adj <= FDR) %>%
  distinct(gene) %>%   
  pull(gene) %>%                        
  na.omit()

# Background Universe Genes 
universe <- df %>%
  distinct(gene) %>% 
  pull(gene) %>%                        
  na.omit()

```


```{r, warning = FALSE}

set.seed(42)

Humanlibrary <- org.Hs.eg.db

###################################### upregulated genes

up.GO <- enrichGO(
  gene          = up_genes,
  universe      = universe,         
  OrgDb         = Humanlibrary,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,       
  minGSSize     = 15,
  maxGSSize     = 800
)

simp.up.GO <- clusterProfiler::simplify(up.GO)

simp.up.GO.show <- as.data.frame(simp.up.GO)
simp.up.GO.show <- simp.up.GO.show %>%
  dplyr::select(-geneID) 
create_table(simp.up.GO.show)

###################################### downregulated genes

down.GO <- enrichGO(
  gene          = down_genes,
  universe      = universe,         
  OrgDb         = Humanlibrary,
  keyType       = "SYMBOL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.1,       
  minGSSize     = 15,
  maxGSSize     = 800
)

simp.down.GO <- clusterProfiler::simplify(down.GO)

simp.down.GO.show <- as.data.frame(simp.down.GO)
simp.down.GO.show <- simp.down.GO.show %>%
  dplyr::select(-geneID) 

create_table(simp.down.GO.show)

```

```{r, warning = FALSE, fig.width=8, fig.height=7}

go_up_df   <- as.data.frame(simp.up.GO) %>% mutate(Direction = "Up-regulated")
go_down_df <- as.data.frame(simp.down.GO) %>% mutate(Direction = "Down-regulated")

go_combined <- rbind(go_up_df, go_down_df) %>%
  group_by(Direction) %>%
  slice_head(n = 10) %>% # Adjust 'n' to show more or fewer terms per panel
  ungroup()


go_plot_data <- go_combined %>%
  mutate(
    GeneRatio_num = sapply(GeneRatio, function(x) {
      num_den <- as.numeric(strsplit(x, "/")[[1]])
      return(num_den[1] / num_den[2])})) %>%
  arrange(Direction, GeneRatio_num) %>%
  mutate(Description = factor(Description, levels = unique(Description)))


ggplot(go_plot_data, aes(x = GeneRatio_num, y = Description)) + 
  geom_point(aes(color = p.adjust, size = Count)) +  
  facet_wrap(~Direction, scales = "free_y", ncol = 1) +
  
  scale_color_viridis_c(
    option = "plasma", 
    name = "adj. p-value",                          
    guide = guide_colorbar(reverse = TRUE)
  ) +
  scale_size_continuous(range = c(3, 8), name = "Gene Count") +
  scale_y_discrete(labels = scales::label_wrap(45)) +
  
  theme_bw() + 
  labs(
    title = "DEG MT CMFHM_TGF vs CMFHM_NC: GO Biological Process Enrichment", 
    subtitle = "Comparing top enriched terms for Up-regulated vs Down-regulated genes",
    x = "Gene Ratio", 
    y = ""
  ) +
  theme(
    plot.title.position = "plot", 
    plot.title    = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle = element_text(size = 10, face = "italic", hjust = 0),
    axis.text.y   = element_text(size = 9, color = "black", face = "bold"),
    axis.title.x  = element_text(face = "bold"),
    strip.text    = element_text(face = "bold", size = 10), 
    legend.position = "right"
  )
```

## EMAP Plot

```{r, warning = FALSE, fig.width=8, fig.height=7}

set.seed(42)

################################################### Upregulated Go terms

simp.up.GO <- pairwise_termsim(simp.up.GO)

emapplot(
  simp.up.GO,
  showCategory = 5,          
  node_label   = "group", 
  nCluster     = 4,           
  nWords       = 4,           
  node_label_size = 3.5       
) + 
  theme_bw() + 
  labs(
    title    = "DEG Fibroblasts CMFHM_TGF vs CMFHM_NC: GO Biological Process Enrichment", 
    subtitle = "Functional clustering map of all enriched terms for Up-regulated genes",
    x        = "", 
    y        = ""
  ) +
  theme(
    plot.title.position = "plot", 
    plot.title      = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle   = element_text(face = "italic", size = 10, hjust = 0),
    axis.text.y     = element_blank(), 
    axis.text.x     = element_blank(),  
    axis.ticks      = element_blank(),  
    panel.grid      = element_blank(),  
    legend.position = "right"
  )

################################################### Down Go terms

simp.down.GO <- pairwise_termsim(simp.down.GO)

emapplot(
  simp.down.GO,
  showCategory = 30,          
  node_label   = "group", 
  nCluster     = 5,           
  nWords       = 4,           
  node_label_size = 3.5       
) + 
  theme_bw() + 
  labs(
    title    = "DEG Fibroblasts CMFHM_TGF vs CMFHM_NC: GO Biological Process Enrichment", 
    subtitle = "Functional clustering map of top 30 enriched terms for Down-regulated genes",
    x        = "", 
    y        = ""
  ) +
  theme(
    plot.title.position = "plot", 
    plot.title      = element_text(face = "bold", size = 12, hjust = 0),
    plot.subtitle   = element_text(face = "italic", size = 10, hjust = 0),
    axis.text.y     = element_blank(), 
    axis.text.x     = element_blank(),  
    axis.ticks      = element_blank(),  
    panel.grid      = element_blank(),  
    legend.position = "right"
  )

```

# Deeper look into specific Genesets 

# Upregulated in Fibroblasts upon TGF-Beta stimulation (With Macrophages)

## extracellular matrix organization

The geneset is quite big so for visualization purposes it will be split in sets of 9

```{r, warning = FALSE, fig.width=8, fig.height=4}

simp.up.GO.df <- as.data.frame(simp.up.GO)

descr <- "extracellular matrix organization"

genes <- simp.up.GO.df %>%
  filter(Description == descr) %>%
  pull(geneID) %>%
  stringr::str_split("/") %>%
  unlist()

SCpubr::do_DotPlot(Fibroblasts, 
                   features = genes,
                   dot.scale = 10,
                   plot.title = "ECM Gene Set Expression Across Conditions") 


```

```{r, warning = FALSE, fig.width=8, fig.height=7}


chunk_size <- 9
total_genes <- length(genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
    plot_subtitle = paste0("Geneset = extracellular matrix organization - Slice ", slice_num),
    seurat_obj    = Fibroblasts,
    colors        = gradient_colors4
  ))
}

```


## tissue development

```{r, warning = FALSE, fig.width=8, fig.height=4}

simp.up.GO.df <- as.data.frame(simp.up.GO)

descr <- "tissue development"

genes <- simp.up.GO.df %>%
  filter(Description == descr) %>%
  pull(geneID) %>%
  stringr::str_split("/") %>%
  unlist()

SCpubr::do_DotPlot(Fibroblasts, 
                   features = genes,
                   dot.scale = 10,
                   plot.title = "tissue development Gene Set Expression Across Conditions") 


```

```{r, warning = FALSE, fig.width=8, fig.height=7}


chunk_size <- 9
total_genes <- length(genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
    plot_subtitle = paste0("Geneset = tissue development - Slice ", slice_num),
    seurat_obj    = Fibroblasts,
    colors        = gradient_colors4
  ))
}

```

# Downregulated in Fibroblasts upon TGF-Beta stimulation (With Macrophages)

## positive regulation of cell migration

```{r, warning = FALSE, fig.width=8, fig.height=4}

simp.down.GO.df <- as.data.frame(simp.down.GO)

descr <- "positive regulation of cell migration"

genes <- simp.down.GO.df %>%
  filter(Description == descr) %>%
  pull(geneID) %>%
  stringr::str_split("/") %>%
  unlist()

SCpubr::do_DotPlot(Fibroblasts, 
                   features = genes,
                   dot.scale = 10,
                   plot.title = "positive regulation of cell migration Gene Set Expression Across Conditions") 


```

```{r, warning = FALSE, fig.width=8, fig.height=7}


chunk_size <- 9
total_genes <- length(genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
    plot_subtitle = paste0("Geneset = positive regulation of cell migration - Slice ", slice_num),
    seurat_obj    = Fibroblasts,
    colors        = gradient_colors4
  ))
}

```

## cell adhesion

```{r, warning = FALSE, fig.width=8, fig.height=4}

simp.down.GO.df <- as.data.frame(simp.down.GO)

descr <- "cell adhesion"

genes <- simp.down.GO.df %>%
  filter(Description == descr) %>%
  pull(geneID) %>%
  stringr::str_split("/") %>%
  unlist()

SCpubr::do_DotPlot(Fibroblasts, 
                   features = genes,
                   dot.scale = 10,
                   plot.title = "cell adhesion Gene Set Expression Across Conditions") 


```

```{r, warning = FALSE, fig.width=8, fig.height=7}

chunk_size <- 9
total_genes <- length(genes)

for (i in seq(1, total_genes, by = chunk_size)) {
  
  end_idx <- min(i + chunk_size - 1, total_genes)
  slice_num <- ((i - 1) / chunk_size) + 1
  print(raincloud.plots(
    gene_list     = genes[i:end_idx],
    plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
    plot_subtitle = paste0("Geneset = cell adhesion - Slice ", slice_num),
    seurat_obj    = Fibroblasts,
    colors        = gradient_colors4
  ))
}

```

# Checking for specific Marker genes 

## Pro-Fibrotic Genes

```{r, warning = FALSE, fig.width=8, fig.height=4}

genes <- c("ACTA2", "FN1", "COL1A1")

SCpubr::do_DotPlot(Fibroblasts, 
                   features = genes,
                   dot.scale = 10,
                   plot.title = "Pro-Fibrotic Genes: Gene Set Expression Across Conditions") 

```

```{r, warning = FALSE, fig.width=8, fig.height=7}

raincloud.plots(
  gene_list     = genes,
  plot_title    = "Fibroblasts CMFHM_TGF vs. CMFHM_NC: Gene distribution profiles across MT conditions",
  plot_subtitle = "Geneset = Pro-Fibrotic genes",
  seurat_obj = Fibroblasts,
  colors = gradient_colors4
)

```
